RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-01-05 manuscripts/manuscript.Rmd
Tightening up: Introduction

Users of the popular statistical and mathematical computing platform R
[@R] enjoy a wealth of readily installable comparative phylogenetic
methods and tools <!-- needs a citation to support it -->. Exploiting
the opportunities arising from this wealth for complex and integrative
comparative research questions relies on the ability to reuse and
integrate previously generated or published data and metadata. The
resulting expanding data exchange needs of the evolutionary research
community are rapidly outpacing the capabilities of the currently
still most widely used data exchange standards, which were all
developed a decade or more ago. This has resulted in a radiation of
different data representations and exchange standard "flavors" that
are no longer interoperable at the very time when the growth of
available data and methods has made that interoperability most
valuable. In response to the unmet needs for standardized data
exchange in phylogenetics, a modern XML-based exchange standard,
called NeXML, has recently been developed [@Vos_2012], and is
beginning to be adopted. NeXML comprehensively supports current data
exchange needs, is predictably machine-readable, and is forward
compatible. To make the capibilities of NeXML available to R users in
an easy-to-use form, and to lower the hurdles to adoption of the
standard, we present RNeXML, an R package that aims to provide easy
programmatic access to reading and writing NeXML documents, tailored
for the kinds of use-cases that will be common for users and
developers of the wealth of evolutionary, and in particular
comparative analysis methods within the R ecosystem.

The exchange problem for phylogenetic data is particularly acute in
light of the challenges in finding and sharing phylogenetic data
without the otherwise common loss of most data and metadata semantics
[@Drew_2013; @Stoltzfus_2012], or with semantics that is predictably
machine interpretable [@Cranston_2014]. For example, the still popular
NEXUS file format [@Maddison_1997] cannot consistently represent
horizontal gene transfer or ambiguity in reading a character (such as
a DNA sequence base pair).  This has led to modifications of NEXUS in
different ways for different needs, with the unfortunate result that
NEXUS files generated by one program can be incompatible with another
[@Vos_2012]. Without a formal grammar, software based on NEXUS files
may also make inconsistent assumptions about tokens, quoting, or
element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
files in the CIPRES portal contain unrecoverable but hard to diagnose
errors.

>>>>

Users of the popular statistical and mathematical computing platform R
[@R] enjoy a wealth of readily installable comparative phylogenetic
methods and tools <!-- needs a citation to support it -->. Exploiting
the opportunities arising from this wealth for complex and integrative
comparative research questions relies on the ability to reuse and
integrate previously generated or published data and metadata. The
expanding data exchange needs of the evolutionary research
community are rapidly outpacing the capabilities of most current
and widely used data exchange standards [@Vos_2012], which were all
developed a decade or more ago. This has resulted in a radiation of
different data representations and exchange standard "flavors" that
are no longer interoperable at the very time when the growth of
available data and methods has made that interoperability most
valuable. In response to the unmet needs for standardized data
exchange in phylogenetics, a modern XML-based exchange standard,
called NeXML, has recently been developed [@Vos_2012]. NeXML
comprehensively supports current data exchange needs, is predictably
machine-readable, and is forward compatible.

The exchange problem for phylogenetic data is particularly acute in
light of the challenges in finding and sharing phylogenetic data
without the otherwise common loss of most data and metadata semantics
[@Drew_2013; @Stoltzfus_2012; @Cranston_2014]. For example, the still popular
NEXUS file format [@Maddison_1997] cannot consistently represent
horizontal gene transfer or ambiguity in reading a character (such as
a DNA sequence base pair).  This and other limitations have led to modifications of NEXUS in
different ways for different needs, with the unfortunate result that
NEXUS files generated by one program can be incompatible with another
[@Vos_2012]. Without a formal grammar, software based on NEXUS files
may also make inconsistent assumptions about tokens, quoting, or
element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
files in the CIPRES portal contain unrecoverable but hard to diagnose
errors.
